Scheduling and Vectorization for MPC

### 2022-03-01 paper_SIMD.tex
Removed todos.

\figref{fig:compiler_framework} presents an overview of our compiler. In this section, we describe several of the phases of the compiler. 
Sections~\secref{sec:vectorization} and~\secref{sec:div_and_conq} describe vectorization and divide-and-conquer.
We write $i,j,k$ to denote the loop nest: $i$ is the outermost loop, $j$, is immediately nested in $i$, and so on until $k$. 
We write $A[i,j,k]$ to denote canonical access to an array element. \ana{Need to state this more precisely.}

@@

\ana{TODO: Still work in progress, all over.}

@@

We define the $\mathit{raise\_dim}$ (raise dimensions) and $\mathit{drop\_dim}$ (drop dimension) functions. Raise dimension "lifts" a lower-dimension array (\ana{The right term here is tensor, I am pretty sure!}) into a higher dimension one. This is necessary when a lower-dimensional array is used in a higher dimensional loop and, essentially, is just copying of values. For example, Biometric contains the statement $d = \mathit{SUB}(S[i, j], C[j])$, in the $j$-loop which is nested into the $i$-loop. Here $C$ is a one-dimensional array, however, to vectorize across both loops it is necessary to turn it into a two-dimensional array: $C[i,j]$ becomes $[C[0],C[1],...C[J],C[0],C[1],...C[J], ... C[0],C[1],...C[J]]$, which turns the row into a matrix of $I$ identical rows. (We use capital letters to denote the upper bounds of loops, e.g., $J$ is the upper bound of the $j$-loop and $I$ is the upper bound of the $i$-loop.)

@@

\ana{This section is still informal. Needs work to make more precise.}

>>>>

\figref{fig:compiler_framework} presents an overview of our compiler. In this section, we describe several of the phases of the compiler. 
Sections~\secref{sec:vectorization} and~\secref{sec:div_and_conq} describe vectorization and divide-and-conquer.
We write $i,j,k$ to denote the loop nest: $i$ is the outermost loop, $j$, is immediately nested in $i$, and so on until $k$
and we use $I,J,K$ to denote the corresponding upper bounds. For simplicity, we write $A[i,j,k]$ to denote canonical access 
to an array element. In the program, canonical access is achieved via the standard row-major order formula: $(J*K)*i + K*j + k$.
To simplify the presentation we describe our algorithms in terms of three-element tuples $i,j,k$.
All discussion generalizes to arbitrarily large loop nests.
%\ana{Need to state this more precisely.}

@@

An important component of our algorithm is the "lifting" of scalars to the corresponding loop dimensionality. 
For example, \texttt{d = S[i * D + j] - C[j]} equiv. to \texttt{d = S[i,j] - C[j]}, which gave rise to \texttt{N*D} subtraction operations in the sequential schedule, 
is lifted. The argument arrays \texttt{S} and \texttt{C} are lifted and the scaler \texttt{d} is lifted: \texttt{d[i,j] = S[i,j] - C[i,j]}.
The algorithm then detects that the statement can be vectorized.

There are three kinds of arrays (for now all kept internally as one-dimensional arrays, but that's under discussion).
\begin{itemize}

\item Scalars: These are scalar variables we lift into arrays for the purposes of vectorization. 
For those, all writes are canonical writes and all reads are canonical reads. We will apply raise dimension when 
a scalar is used in an inner loop (e.g., \texttt{sum0} in line 6 of the MPC source code will be raised to a 1-dimensional 
array since \texttt{sum0} is used in the inner $j$-loop). Drop dimension applies as well; this happens when a scalar
written in an inner loop is used outside of the loop (e.g., \texttt{sum0} for which the lifted inner loop computes 
\texttt{D} values, but the outer loop only needs the last one.)

\item Read-only input arrays: Read-only inputs. There are NO writes, while we may have non-canonical reads, $f(i,j,k)$. 
Phase 1 of Basic vectorization will add raise dimension operation at the beginning of the function to lift these arrays, 
and raise dimension may \emph{reshape} arrays. If there are multiple "views" of the input array, there would be multiple raise dimension statements to create
each one of these views. The invariant is that at reads in loops, the reads of "views" of the original input array are canonical.
Only raise dimension applies to these arrays, and only in the beginning of the program. For example, 1-dimensional array 
\texttt{C} is lifted into 2-dimensional array \texttt{N x D} by copying the row \texttt{N} times.

 \item Read-write output arrays: Writes are canonical (by restriction) but reads can be non-canonical. 
 Dependence analysis limits vectorization when non-canonical read access refers to array writes in
 previous iterations, thus creating loop-carried dependences. We may apply both raise and drop dimension, however, 
 they respect the fixed dimensionality of the output array. The array cannot be raised to a dimension lower than
 its canonical (fixed) dimensionality and it cannot be dropped lower. In addition, non-canonical reads may 
 require lifting (i.e., reshaping) of the array after the most recent write, rather than in the beginning of the program, 
 in order to reduce a non-canonical read to a canonical one.

In the sections below we detail the $\mathit{raise\_dim}$ (raise dimension) and $\mathit{drop\_dim}$ (drop dimension) operations,
followed by our vectorization algorithm.

@@

There are two conceptual versions of $\mathit{raise\_dim}$. One applies on read-only input arrays and reshapes those arrays 
when necessary to ensure canonical read access in the corresponding loop. The signature of $\mathit{raise\_dim}$
is as follows. It takes the original array \texttt{C}, the access pattern function $f(i,j,k)$ in loop nest $i,j,k$ and the loop bounds
\texttt{((i:I),(j:J),(k:K))}: 
\[ \mathit{raise\_dim}(\texttt{A}, f(i,j,k), \texttt{((i:I),(j:J),(k:K))}) \]
It produces a new 3-dimensional array \texttt{A'} by iterating over $i,j,k$ and setting each element of  \texttt{A'} as follows:
\[ \texttt{A'[i,j,k] =  A[f(i,j,k)]} \]
The end result is that uses of \texttt{[A[f(i,jk)]} in loop nest $i,j,k$ are replaced with canonical read-accesses to \texttt{A'[i,j,k]}
that can be vectorized. In the running Biometric example, \texttt{C'} = $\mathit{raise\_dim}$\texttt{(C, j, (i:N,j:D))} lifts the 
1-dimensional array \texttt{C} into a 2-dimensional array. The $i,j$ loop now accesses \texttt{C'} in the canonical way, \texttt{C'[i,j]}. 
Similarly, \texttt{S'} = $\mathit{raise\_dim}$\texttt{(S, i*D+j, (i:N,j:D))} tries to lift \texttt{S}, but the operation turns into a no-op 
because \texttt{S} is already a 2-dimensional array and the read access is canonical.

The other version of $\mathit{raise\_dim}$ applies on scalars and read-write arrays. It lifts a lower-dimension array into a 
higher-dimension for access in a nested loop. Here \texttt{A} is an $i$ array and raise dimension adds two additional dimensions:
\[ \mathit{raise\_dim}(\texttt{A}, \texttt{(j:J,k:K)}) \]
This version is reduced to the above version by adding the access pattern function, which is just $i$: 
\[ \mathit{raise\_dim}(\texttt{A}, i, \texttt{(j:J,k:K)}) \]

The corresponding $\mathit{drop\_dim}$ is carried out when an array written in an inner loop is used in an enclosing loop. 
It takes a higher dimensional array, say $i,j,k$ and removes trailing dimensions, say $j,k$:
\[ \mathit{drop\_dim}(\texttt{A}, \texttt{(j:J,k:K)}) \]
It iterates over $i$ and takes the result at the maximal index of $j$ and $k$, i.e., the result at the last iterations of $j$ and $k$:
\[ \texttt{A'[i] = A[i,J-1,K-1]} \]

\begin{comment}
We define the $\mathit{raise\_dim}$ (raise dimensions) and $\mathit{drop\_dim}$ (drop dimension) functions. Raise dimension "lifts" a lower-dimension array (\ana{The right term here is tensor, I am pretty sure!}) into a higher dimension one. This is necessary when a lower-dimensional array is used in a higher dimensional loop and, essentially, is just copying of values. For example, Biometric contains the statement $d = \mathit{SUB}(S[i, j], C[j])$, in the $j$-loop which is nested into the $i$-loop. Here $C$ is a one-dimensional array, however, to vectorize across both loops it is necessary to turn it into a two-dimensional array: $C[i,j]$ becomes $[C[0],C[1],...C[J],C[0],C[1],...C[J], ... C[0],C[1],...C[J]]$, which turns the row into a matrix of $I$ identical rows. (We use capital letters to denote the upper bounds of loops, e.g., $J$ is the upper bound of the $j$-loop and $I$ is the upper bound of the $i$-loop.)

@@

\ana{This section is still informal. Needs work to make more precise.}
\end{comment}
